Detecting Treatment-Subgroup Interactions in Clustered Data with Generalized Linear Mixed-Effects Model Trees

### 2016-05-19 pkg/glmertree/paper/treatment_interactions.Rnw
- Included discussion of Sies & Van Mechelen paper in paper - Corrected some errors in paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@855 255f4249-2263-4417-b1ea-8498779e27a7

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several tree-based algorithms for the detection of treatment-subgroup interactions have been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}, see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible method for estimating so-called optimal treatment regimes, which allows users to select from a range of statistical methods, including trees, to derive the optimal regimes. As such, the \citeA{ZhanyTsia12Stat} method also provides a tree-based method for treatment-subgroup interaction detection.

Often, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model. For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}. In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}. In tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}. However, none of the purely tree-based methods for treatment-subgroup interaction detection, while also taking account the clustered structure of a dataset. Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models.

The new algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}). MOB offers a flexible framework for subgroup detection, and when the partitioning is based on a generalized linear model (GLM), it allows for detection of treatment-subgroup interactions. Earlier, GLM-based MOB has been applied to detect treatment-subgroup interactions in the treatment of depression \citeA{DrieySmit14} and amyotrophic lateral sclerosis \cite<ALS;>{SeibyZeil15}. The new algorithm we present combines MOB with random-effects estimates and therefore accounts for the clustered structure \cite<which is not done in other purely tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}, as well as treatment effect estimation in models with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.

In what follows, we will discuss existing frameworks the new algorithm builds on: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM). Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees. Subsequently, in the Empirical Evaluation section, we evaluate the performance of the GLMM tree algorithm in simulated datasets. In the Application section, we provide an illustration by applying the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citeA{CuijyWeit14}. But first, we introduce an artificial motivating data set with which will be used to illustrate the methods and algorithms.

@@

As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects. Therefore, we propose the GLMM tree, which combines the GLMM from Equation~\ref{eq:mixedeffects} with the tree from Equation~\ref{eq:fixedeffects_MOB}: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
To estimate the parameters of this model, we take an approach similar to that of \citeA{HajjyBell11} and \citeA{SelaySimo12} but extend their ideas from classical CART trees with only random intercepts to a full GLMM algorithm. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the regression tree, and (2) assuming the regression tree known, allowing for estimation of the random effects.

@@

As discussed in the Introduction, treatment-subgroup interactions have gained interest from several areas, and several different tree-based methods for their detection have been developed. In the current study, we have limited our comparison to the performance of GLM and GLMM-based MOB. Obviously, further research on what the different methods for treatment-subgroup interaction detection have in common and what sets them apart is needed.

>>>>

Tree-based methods are such hypothesis-generating methods, as they can automatically detect subgroups which differ in the expected outcomes for one or more treatments. Due to their flexibility, tree-based methods are preeminently suited to the detection of treatment-subgroup interactions: they can handle many potential predictor variables at once and can automatically detect (higher order) interactions between predictor variables \cite{StroyMall09}. Several tree-based algorithms to the detection of treatment-subgroup interactions have been developed (\citeNP{DussyMech14, DussyMeul04, SuyTsai09, FostyTayl11, LipkyDmit11, ZeilyHoth08}, see \citeNP{DoovyDuss14} for an overview). Also, \citeA{ZhanyTsia12Bio,ZhanyTsia12Stat} have developed a flexible classification-based approach, which allows users to select from a range of statistical methods, including trees. As such, the \citeA{ZhanyTsia12Stat} method can also be referred to as a tree-based method.

Often, researchers may want to detect treatment-subgroup interactions in a generalized linear mixed-effects (GLMM) type model. For example, in individual-level patient data meta-analysis, where datasets of multiple clinical trials on the same treatments are pooled \cite<e.g., >{KoopyHeij07}. In such analyses, the nested or clustered structure of the dataset should be taken into account by including study-specific random effects in the model, prompting the need for a mixed-effects model \cite<e.g., >{CoopyPatt09,HiggyWhit01}. In linear models, ignoring the clustered structure may lead, for example, to biased inference due to underestimated standard errors in linear models \cite<e.g., >{BrykyRaud92,NooryOpde05}. In tree-based methods, ignoring the clustered structure has been found to result in the detection of spurious subgroups and inaccurate predictor variable selection \cite<e.g., >{SelaySimo12, Mart15}. However, none of the purely tree-based methods for treatment-subgroup interaction detection, also take into account the clustered structure of a dataset. Therefore, in the current paper, we present a tree-based algorithm which can be used for the detection of (treatment-subgroup) interactions and non-linearities in GLMM type models.

The new algorithm builds on model-based recursive partitioning (MOB, \citeNP{ZeilyHoth08}). MOB offers a flexible framework for subgroup detection, and when the partitioning is based on a generalized linear model (GLM), it allows for detection of treatment-subgroup interactions. Earlier, GLM-based MOB has been applied to detect treatment-subgroup interactions in the treatment of depression \citeA{DrieySmit16} and amyotrophic lateral sclerosis \cite<ALS;>{SeibyZeil15}. The new algorithm we present combines MOB with random-effects estimation and therefore accounts for the clustered structure \cite<which is not done in other purely tree-based treatment-subgroup interaction detection methods, e.g., >{ZeilyHoth08, SuyTsai09, DussyMech14}, as well as treatment effect estimation in models with continuous and non-continuous response variables \cite<which is not available in previously suggested regression trees with random effects, e.g., >{HajjyBell11,SelaySimo12}.

In what follows, we will discuss existing frameworks the new algorithm builds on: the generalized linear model (GLM), model-based recursive partitioning (MOB), and the generalized linear mixed-effects model (GLMM). Then, we introduce our new algorithm, which combines MOB and the GLMM: generalized linear mixed-effects model trees. Subsequently, in the Empirical Evaluation section, we evaluate the performance of the GLMM tree algorithm in simulated datasets. In the Application section, we provide an illustration by applying the algorithm to an existing dataset of a patient-level meta-analysis on the effects of psycho- and pharmacotherapeutic treatments for depression of \citeA{CuijyWeit14}. But first, we introduce an artificial motivating data set which will be used to illustrate the methods and algorithms.

@@

As noted earlier, ordinary GLM(M)s are not well suited for the detection of treatment-subgroup interactions, whereas the MOB algorithm is, but does not allow for estimation of random effects. Therefore, we propose the GLMM tree, which combines the GLMM from Equation~\ref{eq:mixedeffects} with the tree from Equation~\ref{eq:fixedeffects_MOB}: 
%
\begin{equation}
	\label{eq:glimmertree}
	g(\mu_{i}) = x_{i}^{\top}\beta_{j} + z_{i}^{\top}b
\end{equation}
%
To estimate the parameters of this model, we take an approach similar to that of the mixed-effects regression trees (MERT) approach of \citeA{HajjyBell11} and \citeA{SelaySimo12}, but extend their ideas from classical CART trees with only random intercepts to a full GLMM algorithm. In the MERT approach, the fixed-effects part of a GLMM is replaced by a CART regression tree, and the random-effects part is estimated as usual. To estimate a MERT, an iterative approach is taken, alternating between (1) assuming random effects known, allowing for estimation of the regression tree, and (2) assuming the regression tree known, allowing for estimation of the random effects.

@@

As discussed in the Introduction, several tree-based methods for treatment-subgroup interaction detection are available. These methods have different objectives, and there is not yet an agreed-upon single best method. In a simulation study, \citeA{SiesyVanM16} found the method of \cite{ZhanyTsia12Stat} to perform best, followed by model-based recursive partitioning. However, the method of Zhang et al. performed worst under some conditions of the simulation study in terms of the Type I error rate. Further research comparing tree-based methods for treatment-subgroup interaction detection is needed, especially  for clustered datasets, as our simulations were limited to GLM and GLMM-based MOB.
